---
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(knitr)
library(kableExtra)
```

<div class="main-container">

<div style="background-color:#f5f9ff; border-left:4px solid #2c3e50; padding:20px 24px; margin-bottom:30px; border-radius:4px;">

## Overview of the Methods

<p style="margin-top:10px;">
This project was built to be <strong>transparent, reproducible, and easy to follow</strong>.  
Every graph, table, and comparison on this website comes from a clear, step-by-step workflow in R that someone else could recreate.
</p>

</div>

---

<div style="background-color:#fefaf5; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #f0e0c5;">

## Workflow at a Glance

<p style="margin-top:8px; margin-bottom:10px;">
Most of the analysis on this site follows the same basic pipeline:
</p>

<pre style="background-color:#ffffff; padding:12px 14px; border-radius:4px; border:1px solid #e0dfdf; font-size:0.95em;">
Raw data (NOAA, CDC) 
    ↓
Cleaning & standardizing in R (tidyverse)
    ↓
Joining climate + disease by year/state
    ↓
Summarizing & visualizing (ggplot2, tables)
    ↓
Interpreting patterns & limits
</pre>

<p style="margin-top:10px;">
The goal is not to build the most complex model possible, but to use a clear, repeatable process that makes the climate–disease story visible and honest.
</p>

</div>

---

<div style="background-color:#fcfcfc; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #eeeeee;">

## Data Sources

To explore how climate conditions relate to mosquito-borne disease, this project draws from two major groups of datasets.

### <strong>Climate Data</strong>
- Sourced from <strong>NOAA</strong> climate summaries.  
- Includes <strong>temperature</strong>, <strong>precipitation</strong>, and <strong>drought indices</strong> such as PDSI.  
- Aggregated to the <strong>state–year</strong> level for consistent national comparison.

### <strong>Disease Data</strong>
- Built from human case data for vector-borne diseases using <strong>CDC</strong> public health surveillance.  
- Focuses on four main diseases:
  - <strong>West Nile virus (WNV)</strong>  
  - <strong>Lyme disease</strong>  
  - <strong>Dengue</strong>  
  - <strong>Malaria</strong>  
- Includes reported cases per <strong>state</strong> and <strong>year</strong>.  
- Emphasizes long-term, national patterns rather than individual outbreaks.

<p style="margin-top:10px;">
All datasets were used strictly for <strong>educational purposes</strong> in GEOG 3000 – Climate Change in Science &amp; Society.
</p>

</div>

---

<div style="background-color:#f9fbf9; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #e9f0e9;">

## Data Cleaning &amp; Preparation

All data cleaning was performed in R using the <strong>tidyverse</strong>.

Key steps included:

<ul>
  <li>Importing CSVs with <code>readr::read_csv()</code></li>
  <li>Standardizing column names for readability</li>
  <li>Filtering to valid states and meaningful years</li>
  <li>Checking for missing or inconsistent values</li>
  <li>Removing invalid observations (e.g., negative case counts)</li>
  <li>Creating tidy, analysis-ready tables</li>
</ul>

<p style="margin-top:10px;">
One of the biggest lessons here: <strong>real-world data is imperfect</strong>, and careful cleaning is essential for honest analysis.
</p>

</div>

---

<div style="background-color:#fcfcff; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #ececff;">

## Combining Climate &amp; Disease Data

To make meaningful comparisons, the datasets needed to align.

This involved:

<ul>
  <li>Standardizing state names / abbreviations</li>
  <li>Ensuring matching <strong>year</strong> formats across files</li>
  <li>Creating a shared <strong>state–year</strong> joining key</li>
  <li>Using <code>dplyr::left_join()</code> to merge climate indicators with disease case counts</li>
</ul>

The unified dataset supports the visuals throughout:

<ul>
  <li><strong>Climate Trends</strong></li>
  <li><strong>Disease Trends</strong></li>
  <li><strong>Climate–Disease Comparison</strong></li>
  <li><strong>Utah Focus</strong></li>
</ul>

<p style="margin-top:10px;">
The goal was not prediction, but to create a clean structure where <strong>visual patterns</strong> could be explored with confidence.
</p>

</div>

---

<div style="background-color:#ffffff; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #eeeeee;">

## One Example R Workflow

<p style="margin-top:6px;">
This short example shows the kind of R pipeline used across the project: load data, clean it, join climate and disease, and calculate simple summaries.
</p>

<details>
<summary><strong>Show example R code: load, clean, join, summarize</strong></summary>

```{r example-workflow, echo=TRUE, eval=FALSE}
library(tidyverse)

# 1. Load yearly climate and disease data
climate_temp <- readr::read_csv("data/climate_year_temp.csv") |>
  mutate(year = as.integer(year))

master_disease <- readr::read_csv("data/master_disease_comparison.csv") |>
  mutate(
    year  = as.integer(year),
    cases = as.numeric(cases)
  )

# 2. Filter to one disease (West Nile) and national totals
westnile_us <- master_disease |>
  filter(level == "national", disease == "West Nile") |>
  select(year, cases)

# 3. Join climate and disease by year
climate_westnile <- westnile_us |>
  left_join(climate_temp, by = "year")

# 4. Create a simple summary table
climate_westnile |>
  summarize(
    years_in_study   = n(),
    mean_temp_anom   = mean(temp_anomaly, na.rm = TRUE),
    mean_wnv_cases   = mean(cases, na.rm = TRUE)
  )
```

</details>

<p style="margin-top:10px;">
The graphs and models shown on other tabs are built by extending this same pattern:  
<strong>read &rarr; clean &rarr; join &rarr; summarize &rarr; visualize.</strong>
</p>

</div>

---

<div style="background-color:#f8fafc; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #e5ecf3;">

## Visualization &amp; Analysis in R

All graphs were created using <strong>ggplot2</strong>, chosen for its clear, consistent visual style.

The project uses:

<ul>
  <li><strong>Line charts</strong> for long-term climate and disease trends</li>
  <li><strong>Faceted plots</strong> for comparing regions, states, or diseases</li>
  <li><strong>Scatterplots with trend lines</strong> to explore simple climate–disease relationships</li>
  <li>Clean, colorblind-friendly palettes for accessibility</li>
</ul>

This project emphasizes:

<ul>
  <li>Visual storytelling</li>
  <li>Exploratory data analysis</li>
  <li>Clear communication</li>
</ul>

<p style="margin-top:10px;">
Readers should be able to interpret trends without needing advanced statistical training.
For extra support, the <strong>How to Read These Graphs</strong> page walks through the main plot types step by step.
</p>

</div>

---

<div style="background-color:#eef5ff; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #d3dff5;">

## Reproducibility &amp; Workflow

A major priority was ensuring this project was reproducible.

This was supported by:

<ul>
  <li>Building each page in <strong>R Markdown</strong></li>
  <li>Keeping explanations and code together in the same document</li>
  <li>Using self-updating graphics that knit directly from the underlying data</li>
  <li>Hosting the entire project on <strong>GitHub Pages</strong></li>
  <li>Maintaining a clear, folder-based file structure for data and code</li>
</ul>

<p style="margin-top:10px;">
Reproducibility is not just a coding goal — it’s a commitment to scientific clarity and transparency.
</p>

</div>

---

<div style="background-color:#eaf1ff; padding:20px 24px; border-radius:4px; margin-bottom:25px; box-shadow:0 0 0 1px #d3dff5;">

## Analytical Choices &amp; Limitations

This project uses <strong>descriptive and visual analysis</strong> rather than complex predictive modeling.  
This choice keeps the results accessible and avoids overstating what the datasets can support.

Limitations include:

<ul>
  <li>Differences in reporting practices between states</li>
  <li>Possible under-reporting of human vector-borne disease cases</li>
  <li>Occasional gaps or inconsistencies in climate data</li>
  <li>State-level aggregation that cannot capture city-level or neighborhood variation</li>
</ul>

<p style="margin-top:10px;">
These limitations do not prevent meaningful exploration — they simply frame how the trends should be interpreted.
Readers should treat the results as <strong>patterns to think about</strong>, not as final answers.
</p>

</div>

---

<div style="background-color:#e3edf9; padding:20px 24px; border-radius:4px; margin-bottom:30px; box-shadow:0 0 0 1px #c9d8ee;">

## Software &amp; Versioning

To support reproducibility, all work was completed using:

<ul>
  <li><strong>R</strong> (version 4.x)</li>
  <li><strong>RStudio</strong> as the development environment</li>
  <li><strong>tidyverse</strong> for data cleaning and manipulation</li>
  <li><strong>ggplot2</strong> for visualization</li>
  <li><strong>knitr</strong> and <strong>rmarkdown</strong> for rendering the website</li>
</ul>

<p style="margin-top:10px;">
Using consistent software versions helps ensure that graphs and results remain stable over time and can be reproduced by others following the same steps.
</p>

</div>

<p style="font-size:0.9em; color:#555; margin-top:10px;">
Methods summary for GEOG 3000 – Climate Change in Science &amp; Society  
<br>
Utah Valley University
</p>

</div> <!-- end main-container -->
